---
title: "Toxicity assay on whitefly"
author: "Marc Galland"
date: "`r Sys.Date()`"
output:
    html_document:
        number_sections: yes
        toc: yes
        toc_depth: 2  
params:
  datafile: "../../../../results/01.bioassays/wf/20161104_wf_toxicity_assay/results.txt"
  resultDir: "~/SURFdrive/trichome_team/11.analysis/bioassays/wf/20161104_wf_toxicity_2tridecanone/"
---
# Introduction
This script is intended to analyse so-called __toxicity assays__ where insects are treated with a certain dose of a compound and the number of surviving insects is measured after a certain time (24h for instance). 

We want to calculate the probability for a fly to die (Yi=1) for a given dose (Xi= dose value) 

The input table has to be tabulated (columns separated by tabulation)

sample|dose|alive|dead
------|---------|-----|----
Ctrl-1| Control | 25  | 5
Ctrl-2| Control | 22  | 3
Mock-2| Mock    | 20  | 3
0.2-1 | 0.2µM   | 10  | 12

For model fitting, the first condition (here `Control`) will be used as the _Intercept_


# Setup

## Knitr
```{r setup, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=FALSE,fig.align="center")
```

## Load libraries and set result directory
```{r Load librairies, echo=TRUE,results='hide',message=FALSE,warning=FALSE}
#library(reshape2)
#library(plyr)
#library(dplyr)
library(lme4)
#library(bbmle)
#library(formula.tools)
library(gridExtra)
library(ggplot2)
```

```{r creates result directory}
# creates result directory if it doesn't exist (hide warnings that pop out if it exists)
resdir = params$resultDir
dir.create(resdir,showWarnings = F)

# empty result directory
unlink(resdir)  
```

## data import + show first lines of the table
```{r import data, tidy=FALSE}
wf = read.delim(params$datafile,header = T,na.strings = "NA",stringsAsFactors=F)
print.data.frame(head(wf))

# rename columns
colnames(wf) = c("sample","condition","alive","dead")
```


## Creation of a whitefly % of survival melted dataframe 
```{r WF melted dataframe}
# calculate total number of flies
wf$total = rowSums(wf[,c("alive","dead")])

# calculate percentage of surviving flies
wf$percentage = round(wf$alive / wf$total *100,digits = 0)
```

# Boxplots
```{r Boxplots, echo=FALSE}
# get ordering of the dose (increasing concentrations)
conditionOrder = unique(wf$condition)

# condition as factor (and not as number)
#wf$condition = factor(wf$condition)
#levels(wf$condition) = conditionOrder

# make the plot
p <- ggplot(data = wf,aes(x = condition,y = percentage,fill=condition)) +
  geom_boxplot() +
  #geom_boxplot(aes(group = cut_width(condition,width = 1))) +
  stat_summary(fun.y="mean",geom="point",shape=23,size=2,fill="white") +
  theme(axis.text.x = element_text(angle=0,hjust=1,vjust=1)) + 
  theme(plot.title = element_text(size=18)) +
  theme(legend.title=element_blank()) +
  guides(fill=FALSE) +
  labs(x = "Condition",y = "Whitefly survival after 24h (%)") + 
  scale_x_discrete(limits=conditionOrder)

# print the plot in the final document
print(p)

# save plot into a file
ggsave(filename = file.path(resdir,"boxplot_wf_survival.png"),plot = p,width=10,height=5,dpi = 400)
```

# Logistic regression 

## Remove missing values
```{r Fit models}
# remove missing values
wf = na.omit(wf)

# convert to factor and set first condition (should be control)
wf$condition = as.factor(wf$condition)
wf = within(wf,condition <- relevel(condition,ref = conditionOrder[[1]]))
```

## Fit null model
```{r Fit null model}
# intercept only model (no effect of the condition)
fit1 = glm(cbind(dead,total-dead) ~ 1,data = wf,family = binomial(link = logit)) 
```

## Fit condition model
```{r fit complete model}
# model with the condition of the compound tested
fit2 = glm(cbind(dead,total-dead) ~ 1 + condition, data=wf, family = binomial(link = logit)) 
```

## Compare models (Deviance and AIC plots)
AIC is an information criteria. It can be compared across different models.
A minimal AIC value should be looked for.

[http://data.princeton.edu/R/glms.html](From German Rodriguez Princeton University)
> Selection of terms for deletion or inclusion is based on Akaike's information criterion (AIC). 
> R defines AIC as
> –2 maximized log-likelihood + 2 number of parameters

```{r Compare models}
# put all models in a list
fits = list(fit1,fit2)

# gather deviances and plot
deviances = c(deviance(fit1),deviance(fit2))

# plot AICs and dAICs (differences between AICs) 
AICs = cbind.data.frame(
  models = sapply(fits,function(x) {deparse(formula(x))}),
  AIC = sapply(fits,function(x) AIC(x)),
  dAIC = as.vector(bbmle::AICtab(fit1,fit2)$dAIC)
  )

plotAIC = ggplot(AICs,aes(models,AIC)) + 
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle=45,hjust=1)) +
  ggtitle("AIC for each model")

plotdAIC = ggplot(AICs,aes(models,dAIC)) + 
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle=45,hjust=1)) +
  ggtitle("Changes in AIC between models")

grid.arrange(plotAIC,plotdAIC,ncol=2)

ggsave(filename = file.path(resdir,"AIC.png"),plot = plotAIC,width = 7,height = 5,dpi = 400)
ggsave(filename = file.path(resdir,"dAIC.png"),plot = plotAIC,width = 7,height = 5,dpi = 400)
```

## Compare models with ANOVA
Test (with Chi2 test) if the models significantly differ (based on variance explained)
```{r anova}
# Model comparison with Likelihood Ratio Test
anova(fit1,fit2,test = "LRT") # null model compared to condition model
```


```{r extract coefs}
# Extract coefficients from the best model
# extract coefficients
coefs = as.data.frame(summary(fit2)$coefficients)
colnames(coefs)=c("coeff","stderr","zval","pval")
coefs$condition = sub("^condition",replacement = "",x = row.names(coefs))
row.names(coefs) = NULL
coefs[which(coefs$condition == "(Intercept)"),]$condition <- "untreated"
coefs$condition = factor(coefs$condition,levels = conditionOrder)
# convert coefficients to probabilities (exp(B0 + Bi*Xi=1)/[1+exp(B0 + Bi*Xi=1)]
coefs$prob = sapply(coefs$coeff,function(x) {plogis(x)})

# add confidence interval limits to the coefs (95% confidence)
coefs$lower = sapply(confint(fit2)[,1],function(x) {plogis(x)})
coefs$upper = sapply(confint(fit2)[,2],function(x) {plogis(x)})

# which conditions have the significant pvalues?
coefs$text = NA
coefs[which(coefs$pval > 0.01),]$text <- "ns"
coefs[which(coefs$pval < 0.01),]$text <- "***"

# plot by increasing probability to die
coefs = coefs[order(coefs$condition),]
probas = ggplot(coefs,aes(condition,prob,label=text)) + 
  geom_point() +
  geom_errorbar(ymin = coefs$lower,ymax=coefs$upper,width=0.2) +
  labs(y = "Whitefly death probability",x="Dose") +
  # add pvalues
  geom_text(aes(x = condition,y = upper+0.04*upper),size=5) +  
  guides(fill=FALSE) +
  theme(axis.text.x = element_text(angle=0,hjust=1)) +
  theme(legend.title=element_blank()) +
  scale_y_continuous(limits=c(0,1))
  
# print and save plot
print(probas)
ggsave(filename = file.path(resdir,"proba2die.png"),width = 7,height = 5,dpi = 500)
```

# Session info
```{r}
sessionInfo()
```

